---
title: "Mouse genotype composition analysis"
author: "Mengyuan Shen"
format:
  html:
    theme: minty
    df-print: paged
    code-line-numbers: true
    embed-resources: true
    code-fold: true
    code-fold-default: true
    toc: true
    toc-location: right 
    page-layout: full
knitr:
  opts_chunk:
    message: false
    warning: false
comments:
  hypothesis:
    theme: clean
editor: source
---

# Introduction

This report builds on the mouse thymic project, following quality control report (mouse_thymic_quality_report-2026-01-22.html) and the agreed cell type clusters.

-   Aim: Test whether the composition of TEC clusters differs across genotypes, and characterise transcriptional differences in apoptotic pathways across mimetic populations. This is addressed through compositional modelling (sccomp), pseudobulk gene expression boxplots stratified by genotype and cell type, and summary heatmaps.\
-   Genotypes and hypotheses:
    -   WT: baseline.
    -   BXBN knockout: apoptosis blocked (expected increase in mimetic cells)\
    -   BU knockout: pro-survival lost (expected decrease in mimetic cells)

# Set up

```{r}
library(targets)
library(tidySingleCellExperiment)
library(purrr)
library(dplyr)
library(stringr)
library(Seurat)
library(HPCell)
library(scater) 
library(clustree)
library(readxl)
library(tibble)
library(ggrepel)
if (!requireNamespace("sccomp", quietly = TRUE)) {
  devtools::install_github("MangiolaLaboratory/sccomp")
}
library(sccomp)
library(tidybulk)
source("/vast/projects/thymic_survive/results/utils.R")
integrated_samples <-  readRDS("/vast/projects/thymic_survive/results/filtered/integrated_samples_final_clusters_se.rds") 
integrated_samples = integrated_samples |> mutate(group = ifelse(group == "BU?", "WT", group)) # BU?56 is BCLXL wild type

md <- integrated_samples[[]] |>
  tibble::rownames_to_column("cell") |>
  mutate(genotype = factor(group, levels =  c("WT","BXBN","BU"))) |> 
  dplyr::rename(cell_group = cell_type)
```

# UMAP for each genotype
This section shows all the clusters for each of the genotype.
```{r fig.width=8, fig.height=12}
p <- DimPlot(integrated_samples, group.by = "cell_type", order = T, cols = celltype_colors, split.by = "group", ncol=1 ) 

coords <- p$data |>
  group_by(cell_type) |>
  summarise(
    x = median(umap_1),
    y = median(umap_2)
  )

p +
  geom_text_repel(
    data = coords,
    aes(x = x, y = y, label = cell_type),
    size = 3,
    min.segment.length = 0,    # allow short or no lines
    segment.color = "grey50",
    force = 2,                 # how strongly labels repel each other
    box.padding = 0.5,
    point.padding = 0.3,
    nudge_y = 0.2
  )
```

# Summary

## Samples per genotype

This data frame confirms how many samples contribute to each genotype.

```{r}
integrated_samples |> dplyr::count(Sample, group) |> arrange(group) |> DT::datatable(filter = "top")
```

## Age and sex distribution

```{r}
md |> 
  distinct(Sample, group, Gender, Age) |> 
  ggplot(aes(x = factor(Age), fill = Gender)) +
  geom_bar(position = "dodge") +
  facet_wrap(~ group, nrow = 1, scales = "free_x") +
  theme_bw() +
  labs(
    x = "Mouse Age",
    y = "Number of samples",
    fill = "Gender"
  )

```

## Build composition table

```{r}
# convert per-cell metadata into a **sample × cell_type** count table required by `sccomp`.
sample_meta <- md |>
    distinct(Sample, group, Age, Gender)

df_sccomp <- md |> dplyr::count(Sample, cell_group, name = "cell_count") |> 
  left_join(sample_meta, by = "Sample") |> 
  mutate(group = factor(group, levels = c("WT", "BXBN", "BU")),
         Gender = factor(Gender))

df_sccomp |> DT::datatable(filter = "top")

```

# Composition modeling

We fitted compositional models that estimates how cluster proportions change with genotype while adjusting for phenotypic covariates.\
From discussion, we would typically assume that adult genetically inbred mice within \~1mo of age of each other would be similar enough. For gender, we don’t have a particular reason to assume there will be gender effects for these parameters (there might be! we just don’t know there are). To assess whether gender has an effect, we fitted two models: Model A included genotype only, whereas Model B included gender only.

## Model A:

# $$ \sim genotype$$

```{r}
res_comp <- sccomp_estimate(
  .data = df_sccomp,
  formula_composition = ~ group,
  .sample = Sample,
  .cell_group = cell_group,
  .count = cell_count,
  verbose = F
) |> 
  
  # identify outliers probabilistically based on the model fit, and exlcude them from the estimation
  sccomp_remove_outliers(cores = 1, verbose = FALSE, max_sampling_iterations = 2000) |> 
  sccomp_test()

p <- res_comp |>
  sccomp_boxplot(factor = "group",
                 remove_unwanted_effects = TRUE)
```

## Visualisation and summary plot

Black boxplots show observed within-sample proportions; blue boxplots show posterior predictive distributions from the fitted model. If the model is descriptively adequate for the data, the blue boxplots should roughly overlay the black boxplots, which represent the observed data. The highlighted color means that whether genotype BXBN, and/or BU are significantly different from WT, while transparent box means that there is no significance from WT.

For example, in the AIRE population, WT shows the lowest within-sample proportion and is used as the baseline for comparison. The AIRE proportion is higher in BXBN than in WT, with the transparent box indicating that the difference from WT is not statistically significant. As a comparison, the AIRE proportion is higher in BU than in WT, and this difference is statistically significant. In both BXBN and BU genotypes, the blue posterior predictive boxplots closely overlap the black observed boxplots, indicating that the model adequately captures the observed and predicted distributions for the AIRE population.

```{r fig.width=10, fig.height=8}
p
```

When credible interval is great than 0, cluster is estimated to be more abundant than WT. When credible interval is less than 0, cluster is estimated to be less abundant than WT. Interval crosses 0 means that no strong evidence pf a change vs WT. For example, in BU mice, AIRE cells are credibly increased in proportion compared to the reference genotype.

```{r fig.width=10, fig.height=8}
res_comp |> 
  plot_1D_intervals()
```

## Model B:

# $$ \sim gender$$

```{r}
res_comp_gender <- sccomp_estimate(
  .data = df_sccomp,
  formula_composition = ~ Gender,
  .sample = Sample,
  .cell_group = cell_group,
  .count = cell_count,
  verbose = F
) |> 
  
  # identify outliers probabilistically based on the model fit, and exlcude them from the estimation
  sccomp_remove_outliers(cores = 1, verbose = FALSE, max_sampling_iterations = 2000) |> 
  sccomp_test()

p <- res_comp_gender |>
  sccomp_boxplot(factor = "Gender",
                 remove_unwanted_effects = TRUE)

```

```{r fig.width=10, fig.height=8}
p
```

Since sex is not considered as a parameter, and gender is not significant in BXBN vs WT and BCL vs WT, only group (genotype) parameter is kept in the model.

# Pseudobulk analysis
## Load gene names
```{r}
genes_tbl <- readxl::read_xlsx("/vast/projects/thymic_survive/mimetic TEC RNA-seq apoptotic pathways(Sheet1).xlsx", sheet = 2) |> 
  mutate(Gene_name = stringr::str_trim(Gene_name)) |>
  mutate(Gene_name = case_when(
    Gene_name == "APAF-1"  ~ "Apaf1",
    Gene_name == "RIP1"   ~ "Ripk1",
    TRUE ~ Gene_name
  )) |> 
  mutate(Gene_name = stringr::str_to_title(Gene_name))

map <- AnnotationDbi::mapIds(
      EnsDb.Mmusculus.v79::EnsDb.Mmusculus.v79,
      keys = genes_tbl$Gene_name,
      column = "GENEID",
      keytype = "SYMBOL",
      multiVals = "first"
    ) |>
  enframe("Gene name", "ensemble ID") |>
  mutate(exist_in_object = ifelse(`ensemble ID` %in% rownames(integrated_samples), TRUE, FALSE))
DT::datatable(map, filter = "top")
```

```{r}
# aggregation is based on raw counts rather than sce
obj <- integrated_samples
obj[["RNA"]] <- JoinLayers(obj[["RNA"]])
sce = obj |> as.SingleCellExperiment(assay = "RNA")
ids <- sce |> mutate(aggregated_cells = paste(Sample, cell_type, sep = "___")) |> pull(aggregated_cells)
pb <- scuttle::aggregateAcrossCells(
      sce,
      ids     = ids,
      use.assay.type = "counts",
      BPPARAM = BiocParallel::MulticoreParam(workers = 2),
      store_number = "aggregated_cells"
    ) |> as("SummarizedExperiment")
rn <- rownames(sce)
rownames(pb) <- rn
counts_pb <- SummarizedExperiment::assay(pb, "counts")
logcpm_pb <- edgeR::cpm(counts_pb, log = TRUE, prior.count = 1)
```


# Quantifying apoptotic gene expression across mimetic populations within each genotype

This section compares the transcriptional level of a given apoptotic protein across mimetic populations within the same group (e.g., WT).
```{r}
plot_expr_across_celltypes <- function(expr_long, apotype, grp) {
  expr_long |>
    filter(group == grp, apoptosis_type == apotype) |>
    mutate(
      cell_type = forcats::fct_reorder(cell_type, logCPM, .fun = median, na.rm = TRUE),
      gene_name = forcats::fct_reorder(gene_name, logCPM, .fun = median, na.rm = TRUE)
    ) |>
    ggplot(aes(x = cell_type, y = logCPM)) +
    geom_boxplot(outlier.shape = NA, width = 0.6) +
    geom_point(
      aes(color = Sample, group = Sample),
      position = position_dodge(width = 0.55),
      size = 1.5,
      alpha = 0.8
    ) +
    facet_wrap(~ gene_name, scales = "free_y") +
    theme_bw(base_size = 10) +
    theme(
      plot.title = element_text(size = 20),
      axis.title.x = element_text(size = 15),
      axis.text.x  = element_text(angle = 45, hjust = 1, size = 15),
      axis.title.y = element_text(size = 15),
      strip.text   = element_text(face = "bold", size = 15),
      legend.position = "bottom",
      legend.title = element_text(size = 14),  
      legend.text  = element_text(size = 12)   
    ) +
    labs(
      title = paste0(apotype, " Apoptotic gene expression across mimetic populations (", grp, ")"),
      x     = "Cell type",
      y     = "logCPM (pseudobulk)"
    )
}

plot_expr_across_groups <- function(expr_long, apotype, ct) {
  expr_long |>
    filter(apoptosis_type == apotype, cell_type == ct) |>
    mutate(group = forcats::fct_relevel(group, "WT", "BXBN", "BU")) |>
    ggplot(aes(x = group, y = logCPM, color = group)) +
    geom_boxplot(outlier.shape = NA, width = 0.6) +
    geom_point(
      position = position_dodge(width = 0.55),
      size = 1.5,
      alpha = 0.8
    ) +
    facet_grid(~gene_name, scales = "free_x") +
    theme_bw(base_size = 9) +
    theme(
      plot.title = element_text(size = 15),
      axis.title.x = element_text(size = 10),
      axis.text.x  = element_text(angle = 45, hjust = 1, size = 10),
      axis.title.y = element_text(size = 15),
      strip.text   = element_text(face = "bold", size = 8),
      legend.position = "bottom",
      legend.title = element_text(size = 12),  
      legend.text  = element_text(size = 10)   
    ) +
    labs(
      title = paste0(apotype, " — expression across genotypes in ", ct, " population"),
      x     = "Genotype",
      y     = "logCPM (pseudobulk)",
      color = "Genotype"
    )
}
```

```{r}
# Valid genes present in the expression matrix, with apoptosis type from genes_tbl
map <- map |>
  filter(!is.na(`ensemble ID`), `ensemble ID` %in% rownames(logcpm_pb)) |>
  select(-exist_in_object) |>
  left_join(
    genes_tbl |> select(-name),
    by = c("Gene name" = "Gene_name")
  ) |> distinct()

# Sample metadata
meta_tbl <- pb |>
  colData() |>
  as.data.frame() |>
  transmute(Sample, group, cell_type)

# Long-format tibble: one row per sample × gene
expr_long <- map |>
  purrr::pmap_dfr(function(`Gene name`, `ensemble ID`, apoptosis_type) {
    meta_tbl |>
      mutate(
        gene_name      = `Gene name`,
        ensemble_id    = `ensemble ID`,
        apoptosis_type = apoptosis_type,
        logCPM         = as.numeric(logcpm_pb[`ensemble ID`, ])
      )
  })

DT::datatable(expr_long, filter = "top")
```

## Intrinsic

```{r results='asis', fig.width=20, fig.height=25}
groups <- c("WT", "BXBN", "BU")
cat("::: {.panel-tabset}\n\n")
for (grp in groups) {
  cat(paste0("### ", grp, "\n\n"))
  print(plot_expr_across_celltypes(expr_long, "Intrinsic", grp))
  cat("\n\n")
}
cat(":::\n\n")
```

## Extrinsic

```{r results='asis', fig.width=15, fig.height=12}
groups <- c("WT", "BXBN", "BU")
cat("::: {.panel-tabset}\n\n")
for (grp in groups) {
  cat(paste0("### ", grp, "\n\n"))
  print(plot_expr_across_celltypes(expr_long, "Extrinsic", grp))
  cat("\n\n")
}
cat(":::\n\n")
```

## Non-canonical NF-kB

```{r results='asis', fig.width=15, fig.height=12}
groups <- c("WT", "BXBN", "BU")
cat("::: {.panel-tabset}\n\n")
for (grp in groups) {
  cat(paste0("### ", grp, "\n\n"))
  print(plot_expr_across_celltypes(expr_long, "Non-canonical NF-kB", grp))
  cat("\n\n")
}
cat(":::\n\n")
```


# Quantifying apoptotic gene expression across genotypes within each mimetic population

This section compares the transcriptional level of a given apoptotic protein across groups within the same mimetic population (e.g., BclxL expression in WT vs BclxL KO vs BAX/BAK KO in tuft (1) cells).

## Intrinsic

```{r results='asis', fig.width=10, fig.height=6}
cell_types <- sort(unique(expr_long$cell_type))
cat("::: {.panel-tabset}\n\n")
for (ct in cell_types) {
  cat(paste0("### ", ct, "\n\n"))
  print(plot_expr_across_groups(expr_long, "Intrinsic", ct))
  cat("\n\n")
}
cat(":::\n\n")
```

## Extrinsic

```{r results='asis', fig.width=10, fig.height=6}
cell_types <- sort(unique(expr_long$cell_type))
cat("::: {.panel-tabset}\n\n")
for (ct in cell_types) {
  cat(paste0("### ", ct, "\n\n"))
  print(plot_expr_across_groups(expr_long, "Extrinsic", ct))
  cat("\n\n")
}
cat(":::\n\n")
```

## Non-canonical NF-kB

```{r results='asis', fig.width=10, fig.height=6}
cell_types <- sort(unique(expr_long$cell_type))
cat("::: {.panel-tabset}\n\n")
for (ct in cell_types) {
  cat(paste0("### ", ct, "\n\n"))
  print(plot_expr_across_groups(expr_long, "Non-canonical NF-kB", ct))
  cat("\n\n")
}
cat(":::\n\n")
```

# Heatmap of apoptotic gene expression across mimetic populations

Mean logCPM per cell type × gene is shown, with genes grouped by apoptosis pathway on the Y axis. Each tab shows one genotype.

```{r}
plot_expr_heatmap <- function(expr_long, grp) {
  expr_long |>
    filter(group == grp) |>
    group_by(cell_type, gene_name, apoptosis_type) |>
    summarise(mean_logCPM = mean(logCPM, na.rm = TRUE), .groups = "drop") |>
    mutate(
      cell_type = forcats::fct_reorder(cell_type, mean_logCPM, .fun = mean, na.rm = TRUE),
      gene_name = forcats::fct_reorder(gene_name, mean_logCPM, .fun = mean, na.rm = TRUE)
    ) |>
    ggplot(aes(x = cell_type, y = gene_name, fill = mean_logCPM)) +
    geom_tile(color = "white", linewidth = 0.3) +
    facet_grid(apoptosis_type ~ ., scales = "free_y", space = "free_y") +
    scale_fill_viridis_c(option = "magma", name = "mean\nlogCPM") +
    theme_bw(base_size = 10) +
    theme(
      axis.text.x  = element_text(angle = 45, hjust = 1),
      strip.text.y = element_text(face = "bold", angle = 0),
      panel.spacing = unit(0.3, "lines")
    ) +
    labs(
      title = paste0("Apoptotic gene expression across mimetic populations (", grp, ")"),
      x = "Cell type",
      y = "Gene"
    )
}
```

```{r results='asis', fig.width=10, fig.height=8}
groups <- c("WT", "BXBN", "BU")
cat("::: {.panel-tabset}\n\n")
for (grp in groups) {
  cat(paste0("### ", grp, "\n\n"))
  print(plot_expr_heatmap(expr_long, grp))
  cat("\n\n")
}
cat(":::\n\n")
```

::: {.callout-note collapse="true"}
# Session Info

```{r session-info}
#| label: session-info
sessionInfo()
```
:::
