---
title: "Mouse genotype composition analysis"
author: "Mengyuan Shen"
format:
  html:
    theme: minty
    df-print: paged
    code-line-numbers: true
    embed-resources: true
    code-fold: true
    code-fold-default: true
    toc: true
    toc-location: right 
knitr:
  opts_chunk:
    message: false
    warning: false
comments:
  hypothesis:
    theme: clean
editor: source
---

# Introduction

This report builds on the mouse thymic project, following quality control report (mouse_thymic_quality_report-2026-01-22.html) and the agreed cell type clusters.

-   Aim: Test whether the composition of TEC clusters differs across genotypes.\
-   Genotypes:
    -   WT: baseline.
    -   BXBN knockout: apoptosis blocked (expected increase in mimetic cells)\
    -   BU knockout: pro-survival lost (expected decrease in mimetic cells)

# Set up

```{r}
library(targets)
library(tidySingleCellExperiment)
library(purrr)
library(dplyr)
library(stringr)
library(Seurat)
library(HPCell)
library(scater) 
library(clustree)
library(readxl)
library(tibble)
library(ggrepel)
if (!requireNamespace("sccomp", quietly = TRUE)) {
  devtools::install_github("MangiolaLaboratory/sccomp")
}
library(sccomp)
source("/vast/projects/thymic_survive/results/utils.R")
integrated_samples <-  readRDS("/vast/projects/thymic_survive/results/filtered/integrated_samples_final_clusters_se.rds") 
integrated_samples = integrated_samples |> mutate(group = ifelse(group == "BU?", "WT", group)) # BU?56 is BCLXL wild type

md <- integrated_samples[[]] |>
  tibble::rownames_to_column("cell") |>
  mutate(genotype = factor(group, levels =  c("WT","BXBN","BU"))) |> 
  dplyr::rename(cell_group = cell_type)
```

# Summary

## Samples per genotype

This data frame confirms how many samples contribute to each genotype.

```{r}
integrated_samples |> dplyr::count(Sample, group) |> arrange(group) |> DT::datatable(filter = "top")
```

## Age and sex distribution
```{r}
md |> 
  distinct(Sample, group, Gender, Age) |> 
  ggplot(aes(x = factor(Age), fill = Gender)) +
  geom_bar(position = "dodge") +
  facet_wrap(~ group, nrow = 1, scales = "free_x") +
  theme_bw() +
  labs(
    x = "Mouse Age",
    y = "Number of samples",
    fill = "Gender"
  )

```

## Build composition table

```{r}
# convert per-cell metadata into a **sample × cell_type** count table required by `sccomp`.
sample_meta <- md |>
    distinct(Sample, group, Age, Gender)

df_sccomp <- md |> dplyr::count(Sample, cell_group, name = "cell_count") |> 
  left_join(sample_meta, by = "Sample") |> 
  mutate(group = factor(group, levels = c("WT", "BXBN", "BU")),
         Gender = factor(Gender))

df_sccomp |> DT::datatable(filter = "top")

```

# Composition modeling

We fitted a compositional model that estimates how cluster proportions change with genotype while adjusting for phenotypic covariates.\
From discussion, we would typically assume that adult genetically inbred mice within ~1mo of age of each other would be similar enough. For gender, we don’t have a particular reason to assume there will be gender effects for these parameters (there might be! we just don’t know there are).\
Model:

#  $$ \sim genotype$$


```{r}
res_comp <- sccomp_estimate(
  .data = df_sccomp,
  formula_composition = ~ group,
  .sample = Sample,
  .cell_group = cell_group,
  .count = cell_count,
  verbose = F
) |> 
  
  # identify outliers probabilistically based on the model fit, and exlcude them from the estimation
  sccomp_remove_outliers(cores = 1, verbose = FALSE, max_sampling_iterations = 2000) |> 
  sccomp_test()

p <- res_comp |>
  sccomp_boxplot(factor = "group",
                 remove_unwanted_effects = TRUE)
```

## Visualisation and summary plot
Black boxplots show observed within-sample proportions; blue boxplots show posterior predictive distributions from the fitted model. If the model is descriptively adequate for the data, the blue boxplots should roughly overlay the black boxplots, which represent the observed data. The highlighted color means that whether genotype BXBN, and/or BU are significantly different from WT, while transparent box means that there is no significance from WT.  

For example, in the AIRE population, WT shows the lowest within-sample proportion and is used as the baseline for comparison. The AIRE proportion is higher in BXBN than in WT, with the highlighted boxplot indicating a statistically significant difference from WT. Similarly, the AIRE proportion is higher in BU than in WT, and this difference is also statistically significant. In both BXBN and BU genotypes, the blue posterior predictive boxplots closely overlap the black observed boxplots, indicating that the model adequately captures the observed and predicted distributions for the AIRE population. 
```{r fig.width=10, fig.height=8}
p
```

When credible interval is great than 0, cluster is estimated to be more abundant than WT. When credible interval is less than 0, cluster is estimated to be less abundant than WT. Interval crosses 0 means that no strong evidence pf a change vs WT. For example, in BU mice, AIRE cells are credibly increased in proportion compared to the reference genotype.

```{r fig.width=10, fig.height=8}
res_comp |> 
  plot_1D_intervals()
```

::: {.callout-note collapse="true"}
# Session Info

```{r session-info}
#| label: session-info
sessionInfo()
```
:::
